Phylogenetic adaptions to urban living in Australian birds

### 2018-04-07 ms.Rmd
intro getting pretty flushed out with a reasonable structure, I think.

In order to best predict species' responses to future urbanization changes, it is essential to understand what biological traits are associated with urban birds [e.g., @mcclure1989characterizes; @bonier2007urban; @moller2009successful;  @mcdonnell2015adaptation]. To date, a wide array of traits have been used to assess presence of birds in urban environments. For instance, degree of sociality [@mcclure1989characterizes; @coleman1994neophobia; @jokimaki1998distribution], nesting substrate (CITATION), diet (CITATIONS), distribution size [@jokimaki2000artificial; @chace2006urban; @croci2008does], migratory status [@friesen1995effects], niche breadth [@kark2007living; @evans2011make], and fecundity [@croci2008does; @moller2009successful; @vaugoyeau2016interspecific], among other traits (see Appendix 1), have all been hypothesized to affect a birds' presence in urban environments. However, a wide range of results have been reported in the literature, often contrasting one another. Brain size was found to be important by [@moller2015brain & @maklakov2011brains] but [@evans2011make & @kark2007living] found that brain size had no effect on species' response to urbanization. Similarly, contrasting results have been reported for annual fecundity [cf., @evans2011make with @moller2009successful & @croci2008does] and generalism or niche breadth [cf., @kark2007living with @evans2011make].

The majority of previous studies place species in discrete groups such as urban avoiders, adapters, or exploiters [e.g., @mckinney2002urbanization; @mckinney2006urbanization; @kark2007living; @mcdonnell2015adaptation] or even birds which do or do not occur within urban environments [e.g., @bonier2007urban; @moller2009successful]. This severely limits the generalization of these studies' results as it assumes species within a group respond to urbanization equally [@evans2011make]. Moreover, the terminology (i.e., urban adapters, avoiders, and exploiters) is used differently among studies [cf., [@kark2007living] and [@croci2008does]], potentially affecting interpretation between scientists and the public [@mcdonnell2015adaptation; @fischer2015categorizing]. In a notable exception, [@evans2011make] advocates for a continuous measure of urbanization in which a species' is assessed on their entire distribution in response to urban environments.

Despite the number of studies which have investigated the relationships between species' traits and urban adaptedness, most studies are either spatially limited, such as a single city or region [@blair1996land; @kark2007living; @croci2008does]; or temporally limited to a single breeding season [@blair1996land] or even two month survey period [@kark2007living]]. Most studies also neglect any phylogenetic constraints in their study design [but see @bonier2007urban; @evans2011make], or their analyses are conducted within a single closely-related group of birds [@maklakov2011brains]. We note that the generally small sample sizes (i.e., number of species) might not require the use such phylogenetically-controlled analyses.

Fortunately, citizen science [see reviews by @devictor2010beyond; @tulloch2013realising; @bonney2014next; @kobori2016citizen] can provide broad-scale empirical datasets, allowing us to test hypotheses at spatial and temporal scales previously difficult to attain. This data provides a cost-effective method to research the relationship between species' traits and urban observations. Various projects have used citizen science data to elucidate information on abundance and distribution of birds in urban ecosystems [@mccaffrey2005using; @callaghan2017assessing], as well as functional composition of urban birds [@la2014role; @morelli2016evidence; @lepczyk2017global].

Urban avoiders are sensitive to urban environments [@mcdonnell2015adaptation] and tend to apportion more energy towards reproduction, relative to adapters and exploiters [@croci2008does]. Urban adapters can utilize urban environmental resources but may not reside in them [@mcdonnell2015adaptation] and are generally sedentary, social, high fecundity, with wide distribution [@kark2007living, @croci2008does]. Urban exploiters thrive in urban environments [@mcdonnell2015adaptation] and are generally more social and sedentary than adapters [@kark2007living]. Species’ traits such as social status, behavioral flexibility, migratory status, nesting, diet preferences, habitat preferences, and overall species range have all been used to predict presence in urban environments [@mcclure1989characterizes; @mills1989effects; @jokimaki1998distribution; @lim2004responses; @clergeau2006avifauna; @chace2006urban; @lowry2013behavioural]. Understanding which favored traits are associated with urban adapters and avoiders [@croci2008does] and their habitats can inform management decisions for conservation of particular species and improvements of habitats within urban environments.

Rather than a specific trait that predicts a species’ ability to be an urban avoider, adapter or exploiter, it is more likely a combination of traits. For instance, urban adapters’ presence in Jerusalem was predicted by a combination of traits including diet, degree of sociality, sedentariness, and preferred nesting sites [@kark2007living]. While a regional analysis throughout France and Switzerland demonstrated urban adapters prefer forested habitats, are sedentary, omnivorous, high-nesters, and were widely distributed [@croci2008does].

Here, we use data from eBird [@sullivan2009ebird; -@sullivan2014ebird], a successful citizen science project with over 500 million observations, submitted by more than 250,000 participants [@sullivan2017using], in order to elucidate adaptations in Australian birds to urban living. We combine eBird data with multiple input rasters, that represent a continuous levels of urbanness, over the whole of Australia. This is opposed to assigning species to a priori groups (adapters, avoiders, or exploiters) and investigating differences in traits among groups [e.g., @kark2007living]. We then apply phylogenetically controlled models in order to assess species traits that predict pre-adaptation to urban living. We hypothesize that a combination of traits allows species to thrive in urban environments.

@@

We filtered the bird observations (sensu [@callaghan2017assessing; @la2014role]) by only including observations which were on complete checklists and which followed these protocols: travelling, random, stationary, area, or BirdLife Australia protocols [@barrett2003new]. For further information on protocols see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types]. Further, we filtered the checklists to those with a distance travelled of < 5 km or an area covered which was < 500 Ha. Any checklists which were shared among multiple individuals were subsampled to include only one checklist in order to avoid duplicate results. After this sampling, we were left with 637,482 eligible checklists throughout Australia.

@@

## Response variable - an urbanization index
We used VIIRS night-time lights [@elvidge2017viirs] as a proxy of a bird's association with urban habitat. The VIIRS night-time lights measures the electric lighting present on Earth's surface which is generally associated with human settlement. There are also a significant number of steps (e.g., filtering out background noise, degraded data, and other light source contamination) which are used to ensure the validity of the data and its association with human settlement [@elvidge2017viirs]. For the 637,482 checklists which met our criteria above, we used the checklists' associated spatial coordinates and Google Earth Engine [@gorelick2017google] to assign the average radiance value calculated from the VIIRS layer. For each location, the average radiance was calculated within a 5 km buffer, in order to match the spatial scale of the checklists included and to minimize any potential bias in eBird sampling protocols. Any radiance values which were negative were forced to 0.00001 (sensu CITATION!!!).

@@

For both approaches, we employed a model-averaging technique [@grueber2011multimodel] in which we 'dredged' all possible subsets of the model [@barton2016mumin]. The global model was weighted by a measure of error associated with the response variable. This was assessed as the proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This helps to give more weight in the model-fitting procedure to those species that had more observations in more unique locations (i.e., an assessment of reliability of the observations). The global model fit was assessed by averaging the R2 values [@nakagawa2013general; @nakagawa2017coefficient] for each model in the top-selected model-set. Additionally, we calculated the relative contribution of each explanatory variable [sensu @evans2011make] by **XXXXXXXXXXXXXXXX**. The phylogenetically constrained model was fit using the 'phylolm' function from the phylolm package [@ho2014phylolm]. In order to assess the model fit and relative effect sizes of explanatory variables in the phylogenetically constrained models we **XXXXXXXXXXXX**.

>>>>

In order to best predict species' responses to future urbanization changes, it is essential to understand what biological traits are associated with urban birds [e.g., @mcclure1989characterizes; @bonier2007urban; @moller2009successful;  @mcdonnell2015adaptation]. To date, a wide array of traits have been used to assess presence of birds in urban environments. For instance, degree of sociality [@mcclure1989characterizes; @coleman1994neophobia; @jokimaki1998distribution], nesting substrate [@kark2007living; @croci2008does], diet [@beissinger1982effects; @fuller2008garden; @major2010museum; @evans2011make], distribution size [@jokimaki2000artificial; @chace2006urban; @croci2008does], migratory status [@friesen1995effects], niche breadth [@kark2007living; @evans2011make], and fecundity [@croci2008does; @moller2009successful; @vaugoyeau2016interspecific], among other traits (see Appendix 1), have all been hypothesized to affect a birds' presence in urban environments. Rather than a specific trait that predicts a species’ ability to be found in urban environments, it is most likely a combination of traits. For instance, urban birds' presence in Jerusalem was predicted by a combination of traits including diet, degree of sociality, sedentariness, and preferred nesting sites [@kark2007living], while a regional analysis throughout France and Switzerland demonstrated that urban birds preferred forested habitats, were sedentary, omnivorous, high-nesters, and widely distributed [@croci2008does]. Notably, a wide range of results using the aforementioned traits have been reported in the literature, often contrasting one another. Indeed, brain size was found to be important by [@moller2015brain & @maklakov2011brains] but [@evans2011make & @kark2007living] found that brain size had no effect on species' response to urbanization. Similarly, contrasting results have been reported for annual fecundity [cf., @evans2011make with @moller2009successful & @croci2008does] and generalism or niche breadth [cf., @kark2007living with @evans2011make].

The majority of previous studies place species in discrete groups such as urban avoiders, adapters, or exploiters [e.g., @mckinney2002urbanization; @mckinney2006urbanization; @kark2007living; @mcdonnell2015adaptation] or even birds which do or do not occur within urban environments [e.g., @bonier2007urban; @moller2009successful]. This type of discrete response variable severely limits the generalization of these studies' results as it assumes species within a group respond to urbanization equally [@evans2011make]. Moreover, the terminology (i.e., urban adapters, avoiders, and exploiters) is used differently among studies [cf., [@kark2007living] and [@croci2008does]], potentially affecting interpretation between scientists and the public [@mcdonnell2015adaptation; @fischer2015categorizing]. In a notable exception, [@evans2011make] advocates for a continuous measure of urbanization in which a species' is assessed on their entire distribution in response to urban environments.

Despite the number of studies which have investigated the relationships between species' traits and urban adaptedness, most studies are either spatially limited, such as a single city or region [@blair1996land; @kark2007living; @croci2008does]; or temporally limited to a single breeding season [@blair1996land] or even two month survey period [@kark2007living]]. Fortunately, citizen science [see reviews by @devictor2010beyond; @tulloch2013realising; @bonney2014next; @kobori2016citizen] can provide broad-scale empirical datasets, allowing us to test hypotheses at spatial and temporal scales previously difficult to attain. This data provides a cost-effective method to research the relationship between species' traits and urban observations. Indeed, various projects have used citizen science data to elucidate information on abundance and distribution of birds in urban ecosystems [@mccaffrey2005using; @callaghan2017assessing], as well as functional composition of urban birds [@la2014role; @morelli2016evidence; @lepczyk2017global]. However, we unaware of any projects which use citizen science observations to assess species' traits in relation to urban environments over a large spatial and temporal scale.

Here, we improve upon current methods of assessing biological traits which predicts urbanization in birds by using a series of publicly available datasets available for Australian birds. Importantly, we developed and used a continuous urbanization index as a response variable in the analyses. We had two objectives. Our first objective was to investigate whether there was a phylogenetic signal in our urbanization index, as previous studies have found a lack of phylogenetic signal in urbanization [@evans2011make]. Second, we developed models which were and were not phylogenetically constrained to assess species' traits that predict pre-adaptation to urban living, using continuous and categorical explanatory variables. These results are especially critical in the future action plans of Australia's birds [@garnett2011action], given Australia is currently undergoing a period of rapid urban growth.

@@

We filtered the bird observations [@callaghan2017assessing; @la2014role] by only including observations which were on complete checklists and which followed these protocols: travelling, random, stationary, area, or BirdLife Australia protocols [@barrett2003new]. For further information on protocols see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types]. Further, we filtered the checklists to those with a distance travelled of < 5 km or an area covered which was < 500 Ha. Any checklists which were shared among multiple individuals were subsampled to include only one checklist in order to avoid duplicate results. After this sampling, we were left with 637,482 eligible checklists throughout Australia.

@@

## Response variable - an urbanization index
We used VIIRS night-time lights [@elvidge2017viirs] as a proxy of a bird's association with urban habitat. The VIIRS night-time lights measures the electric lighting present on Earth's surface which is generally associated with human settlement. There are also a significant number of steps (e.g., filtering out background noise, degraded data, and other light source contamination) which are used to ensure the validity of the data and its association with human settlement [@elvidge2017viirs]. For the 637,482 checklists which met our criteria above, we used the checklists' associated spatial coordinates and Google Earth Engine [@gorelick2017google] to assign the average radiance value calculated from the VIIRS layer. For each location, the average radiance was calculated within a 5 km buffer, in order to match the spatial scale of the checklists included and to minimize any potential bias in eBird sampling protocols. Any radiance values which were negative were forced to 0.00001 [*sensu* @ou2015evaluation].

@@

For both approaches, we employed a model-averaging technique [@grueber2011multimodel] in which we 'dredged' all possible subsets of the model [@barton2016mumin]. The global model was weighted by a measure of error associated with the response variable. This was assessed as the proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This helps to give more weight in the model-fitting procedure to those species that had more observations in more unique locations (i.e., an assessment of reliability of the observations). The global model fit was assessed by averaging the R2 values [@nakagawa2013general; @nakagawa2017coefficient] for each model in the top-selected model-set. Additionally, we calculated the relative contribution of each explanatory variable [*sensu* @evans2011make] by **XXXXXXXXXXXXXXXX**. The phylogenetically constrained model was fit using the 'phylolm' function from the phylolm package [@ho2014phylolm]. In order to assess the model fit and relative effect sizes of explanatory variables in the phylogenetically constrained models we **XXXXXXXXXXXX**.
